---
title: "Fire severity - recalcitrant carbon"
output:
  html_document: default
  word_document:
    fig_height: 6
    fig_width: 8
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(here)

library(dplyr)
library(ggplot2)
library(reshape2)
library(stringr)

library(rjags)
load.module("glm")

set.seed(1234)

# short-cut functions for quantiles
lwr <- function(x) quantile(x, probs=0.025)
upr <- function(x) quantile(x, probs=0.975)

# Set to FALSE to run model from scratch or FALSE to 
# load previous model outputs
RUN.MODEL <- FALSE


# ggplot bits
t <- theme_bw() +
  theme(text = element_text(size = 16))

theme_set(t)


load_from <- function(.path) {
  load(.path)
  obs <- ls(all.names = FALSE)
  if (length(obs) == 1) get(obs[1], inherits = FALSE)
  else if (length(obs) > 1) mget(obs, inherits = FALSE)
  else NULL
}

```


## Note on depth categories

For the published paper depth classes are referred to as 0-5cm and 5-15cm. However, in the data and analyses here the second class was labelled as 6-15cm so re-labelling is done when creating graphs with ggplot.

```{r}

DAT <- load_from( here::here("data", "firesev_rpc.RData") )

if (!RUN.MODEL) {
  path <- here::here("model_outputs")

  # Load model predictions from earlier run
  # (if more than 1 file take the last which should be most recent)
  f <- tail( dir(path, pattern="rpc_model_predictions.*RData$", full.names = TRUE), 1)
  M.predicted <- load_from(f)

  # Load model parameter samples
  f <- tail( dir(path, pattern="rpc_model_parameters.*RData$", full.names = TRUE), 1)
  m.sims <- load_from(f)  
}

```

```{r}

# Function which takes an mcmc object generated with coda.samples
# and calculates a summary data frame of rates of inclusion and 
# coefficient values for each predictor.
#
# m - matrix of MCMC samples extracted from model outputs. It is 
#     assumed to have a set of columns for coefficient samples and
#     a matching set of columns for the corresponding binary indicator
#     variables.
#
makeVarSummary <- function(m, indicator.prefix = "^ind_") {
  
  indicator.cols <- which( str_detect(colnames(m), indicator.prefix) )
  var.names <- str_replace(colnames(m)[indicator.cols], indicator.prefix, "")
  var.cols <- match(var.names, colnames(m))
  
  if (anyNA(var.cols)) stop("Unmatched column names")
  
  Nterms <- length(var.cols)
  
  # Rate of variable inclusion
  var.inclusion <- apply(m[, indicator.cols], 2, mean)
  
  # Coefficient values: included and overall
  var.coefs <-
    sapply(1:Nterms, 
           function(i) {
             x <- m[, var.cols[i]]
             ind <- m[, indicator.cols[i]]
             
             x.in <- x[ ind == 1 ]
             if (length(x.in) == 0) c(0, 0, 0, 0, FALSE)
             else {
               qs <- quantile(x.in, probs=c(0.025, 0.975))
               non.zero <- sign(qs[1]) != 0 & sign(qs[1]) == sign(qs[2])
               c(mean(x.in), mean(x), qs, non.zero)
             }
           })
  
  # Create output data frame

  var.summary <- data.frame(
    Variable = var.names,
    PercentInclusion = 100 * var.inclusion,
    MeanIncluded = var.coefs[1, ],
    MeanOverall = var.coefs[2, ],
    Lower = var.coefs[3, ],
    Upper = var.coefs[4, ],
    NonZero = var.coefs[5, ],
    stringsAsFactors = FALSE
  )
  
  arrange(var.summary, desc(PercentInclusion))
}

```



```{r}

# We generate data sets even if not running the model because they
# can be useful for summaries etc


# First create a data frame of digest values where rows are samples 
# (sampleIndex in DAT) and cols are digest reps. Give missing digest 
# reps NA values.
#

yobs <- DAT %>% 
  select(sampleIndex, digestPercentC) %>%
  mutate(logRecalC = log10(digestPercentC)) %>%
  group_by(sampleIndex) %>% 
  mutate(digestRep = row_number()) %>% 
  ungroup() %>%
  dcast(sampleIndex ~ digestRep, value.var = "logRecalC")

# Predictor data thinned to one record per field sample
DAT.thinned <- DAT %>%
  select(sampleIndex, depth, microsite, severity, predigestPercentC) %>%
  distinct(sampleIndex, .keep_all = TRUE) %>%
  mutate(logInitC = log10(predigestPercentC))

# Put predictor data in same order as yobs data frame
DAT.thinned <- left_join(select(yobs, sampleIndex), DAT.thinned)

X.formula <- formula(~ logInitC * depth +
                       depth * severity +
                       microsite * severity)

# Model matrix for regression
X <- model.matrix(X.formula, data = DAT.thinned)
Nterms <- ncol(X)


# Constraint matrix on variable inclusion
X.constraints <- matrix(0, nrow=Nterms, ncol=Nterms)
term.names <- colnames(X)
rownames(X.constraints) <- term.names
colnames(X.constraints) <- term.names

# This recursive function takes a constraint matrix row index
# and a term name and updates the matrix for inclusion of the
# relevant component terms.
#
setConstraint <- function(X.constraints, row.index, term.name) {
  if (is.null(term.name)) {
    term.name <- term.names[row.index]
    term.index <- row.index
  } else {
    term.index <- match(term.name, term.names)
  }
  
  X.constraints[row.index, term.index] <- 1
  
  torder <- 1 + str_count(term.name, "\\:")
  
  if (torder > 1) {
    components <- combn( unlist( str_split(term.name, "\\:")), torder-1 )
    components <- apply(components, 2, paste, collapse=":")
    
    for (component in components) {
      X.constraints <- setConstraint(X.constraints, row.index, component)
    }
  }
  
  X.constraints
}

# Run the function to populate the constraint matrix.
#
for (i in 1:Nterms) {
  X.constraints <- setConstraint(X.constraints, i, NULL)
}


# Data and model matrix for posterior predictions.
# This has one row for each of the 24 possible combinations
# of the factor predictors:
#  2 depths x 3 microsites x 4 fire severity combinations
#
# The initial carbon values are the overall median by depth.
#
DAT.predict <- DAT.thinned %>%
  select(depth, microsite, severity) %>%
  distinct()

Carbon.predict <- DAT.thinned %>%
  group_by(depth, microsite, severity) %>%
  summarize(logInitC = log10( mean(predigestPercentC) ))

DAT.predict <- left_join(DAT.predict, Carbon.predict,
                         by=c("depth", "microsite", "severity"))

X.predict <- model.matrix(X.formula, data = DAT.predict)

```


```{r eval = RUN.MODEL}

modelTxt <- "
  model {

    ############################################################
    # First step - infer 'true' value of recalcitrant carbon for
    # each field sample via t-distributions. 
    ############################################################

    for (i in 1:N) {
      for (j in 1:Ndigest) {
        yobs[i, j] ~ dt(ytrue[i], sd.digest^(-2), nu)
      }
    }

    sd.digest ~ dunif(0, 10)
    
    nuMinusOne ~ dexp(1 / 29.0)
    nu <- nuMinusOne + 1


    ###################################################
    # Second step - relate 'true' values to predictors.
    ###################################################

    for (i in 1:N) { 
      ytrue[i] ~ dnorm(mu[i], sigma^(-2)) 
    }

    # calculation of mean responses
    mu <- X %*% beta

    # Predictor inclusion: first draw unconstrained values
    for (i in 1:Nterms) {
      indDash[i] ~ dbern(p.ind)
    }

    # Predictor inclusion: now apply constraints to ensure
    # interaction terms are accompanied by their component terms
    # (note >0 test to reduce all non-zero terms to 1)
    ind <- (indDash %*% X.constraints) > 0

    # Predictor inclusion: finally apply indicators to 
    # coefficients
    for (i in 1:Nterms) {
      betaDash[i] ~ dnorm(0, sigma.ind^(-2))
      beta[i] <- betaDash[i] * ind[i]
    }

    p.ind ~ dbeta(1, 1)

    # Prior on common standard deviation of field samples
    sigma ~ dunif(0, 10)

    # Prior on standard deviation for variable inclusion
    sigma.ind ~ dunif(0, 10)
    
    ###########################################
    # Posterior predictions
    ###########################################
    
    mu.predict <- X.predict %*% beta
    
    for (i in 1:Npredict) { 
      ypredict[i] ~ dnorm(mu.predict[i], sigma^(-2)) 
    }  
  
  }
"

zz <- textConnection(modelTxt)

model <- jags.model(zz, 
                    
                    data=list(yobs = yobs, 
                              X = X,
                              X.constraints = X.constraints,
                              N=nrow(yobs), 
                              Ndigest=ncol(yobs),
                              Nterms = Nterms,
                              X.predict = X.predict,
                              Npredict = nrow(X.predict)),
                    
                    inits = function(...) {
                      list(
                        ytrue = rep(0, nrow(yobs)),
                        sd.digest = runif(1, 0, 1),
                        nuMinusOne = runif(1, 1, 20),
                        betaDash = runif(Nterms, -1, 1), 
                        sigma = runif(1, 0.1, 10),
                        sigma.ind = runif(1, 0.1, 10),
                        p.ind = runif(1, 0.25, 0.75),
                        indDash = rep(0, Nterms),
                        ypredict = rep(0, nrow(X.predict)) )
                      }, 
                    
                    n.chains = 1)

close(zz)

update(model, 50000)

sims <- coda.samples(model, 
                     c("ytrue", "nu", "sd.digest", "beta", "ind", "p.ind", "ypredict"), 
                     n.iter=100000, 
                     thin=10)


# Extract two matrices from the MCMC output object:
# - matrix of model parameter samples
# - matrix of posterior predictions joined to prediction case data
#
m.sims <- as.matrix(sims[[1]])
rm(sims)
gc()
ii.predict <- str_detect(colnames(m.sims), "^ypredict")
M.predicted <- m.sims[, ii.predict]


# Transpose the matrix of predictions and join it to the
# predict case data
M.predicted <- t(M.predicted)
M.predicted <- cbind(DAT.predict, M.predicted)


# Save this matrix to the outputs dir so that we don't have to
# run the model everytime we've cleared out the workspace.
#
time <- format(Sys.time(), "%Y-%m-%d-%H-%M")
path <- "model_outputs"

save(M.predicted, 
     file = here::here(path, paste0("rpc_model_predictions_", time, ".RData")))

# Separately save the matrix of samples for other variables.
# We change 'beta' and 'ind' column names to names of predictors in X
# to make subsequent use easier, and get rid of pesky square brackets 
# in the remaining column names.

ii.beta <- str_detect(colnames(m.sims), "^beta\\[")
colnames(m.sims)[ii.beta] <- colnames(X)

ii.ind <- str_detect(colnames(m.sims), "^ind\\[")
colnames(m.sims)[ii.ind] <- paste("ind", colnames(X), sep="_")

# purge brackets from column names
colnames(m.sims) <- str_replace_all(colnames(m.sims), "\\[|\\(|\\]|\\)", "")

save(m.sims[, !ii.predict],
     file = here::here(path, paste0("rpc_model_parameters_", time, ".RData")))

```


## Variable importance

```{r}
# Generate summary of percent inclusion and coefficient values for predictors
varimp <- makeVarSummary(m.sims)
knitr::kable(varimp, digits=4)

write.csv(varimp, 
          file = here::here("model_outputs", "rpcarbon_varimp.csv"), 
          row.names = FALSE)

```


```{r}

# Column indices of predicted values in M.predicted
predict.cols <- (ncol(DAT.predict) + 1):ncol(M.predicted)

```


## Model predictions by depth, fire severity

```{r}

dat <- melt(M.predicted, 
            id.vars = c("severity", "depth"), 
            measure.vars = predict.cols) %>%
  mutate(value = 10^value)

ggplot(data = dat) + 
  geom_density(aes(x = value, colour = severity),
               size = 1) +
  
  scale_colour_discrete(name = "Fire severity") +
  scale_x_log10() +
  
  labs(x = "%RPC") +
  facet_wrap(~ depth, ncol=1)


# Tabular summary of the above predictions

x <- dat %>%
  group_by(depth, severity) %>%
  summarize(
    mean = mean(value),
    lwr = quantile(value, 0.025),
    upr = quantile(value, 0.975))

knitr::kable(x, digits=2)

```


## Model predictions by depth, micro-site

```{r}

dat <- melt(M.predicted, 
            id.vars = c("microsite", "depth"), 
            measure.vars = predict.cols) %>%
  mutate(value = 10^value)

ggplot(data = dat) + 
  geom_density(aes(x = value, colour = microsite),
               size = 1) +
  
  scale_colour_discrete(name = "Micro-site") +
  scale_x_log10() +
  
  labs(x = "%RPC") +
  facet_wrap(~ depth, ncol=1)


# Tabular summary of the above predictions

x <- dat %>%
  group_by(depth, microsite) %>%
  summarize(
    mean = mean(value),
    lwr = quantile(value, 0.025),
    upr = quantile(value, 0.975))

knitr::kable(x, digits=2)
```



# Model predictions by depth, micro-site, fire severity

```{r}

dat <- melt(M.predicted, 
            id.vars = c("microsite", "depth", "severity"), 
            measure.vars = predict.cols) %>%
  mutate(value = 10^value)

ggplot(data = dat) + 
  geom_density(aes(x = value, colour = severity),
               size = 1) +
  
  scale_colour_discrete(name = "Fire severity") +
  scale_x_log10() +
  
  labs(x = "%RPC") +
  facet_grid(microsite ~ depth)


# Tabular summary of the above predictions

x <- dat %>%
  group_by(depth, microsite, severity) %>%
  summarize(
    mean = mean(value),
    lwr = quantile(value, 0.025),
    upr = quantile(value, 0.975))

knitr::kable(x, digits=2)

```


## Graph of means and intervals corresponding to above full-breakdown table

This is the graph for the published paper. Note that depth labels have been changed to '0-5cm' and '5-15cm' in this graph to address reviewer comments for the paper.

```{r}

ggplot(data = x) +
  geom_errorbar(aes(x = severity, group = microsite, ymin = lwr, ymax = upr),
                width = 0.5,
                position = position_dodge(width = 0.7)) +
  
  geom_point(aes(x = severity, y = mean, shape = microsite), 
             size = 3,
             position = position_dodge(width = 0.7)) +
  
  scale_shape(name = "micro-site", 
              breaks = c("O", "R", "S"),
              labels = c("open", "rough", "smooth")) +
  
  labs(x = "Fire intensity", y = "%RPC") +
  
  facet_wrap(~ depth, ncol = 1,
             labeller = as_labeller(c('0-5cm' = '0-5cm', '6-15cm' = '5-15cm')))

```


## Fire severity differences within depth and micro-site

```{r}

dat <- melt(M.predicted, 
            id.vars = c("severity", "depth", "microsite"), 
            measure.vars = predict.cols) %>%
  mutate(value = 10^value)

dat <- dcast(dat, 
             depth + microsite + variable ~ severity, 
             value.var = "value")

# calculate pairwise differences in predictions between fire severities
dat <- mutate(dat, 
              LH_LL = LH - LL,
              HL_LL = HL - LL,
              HH_LL = HH - LL,
              HL_LH = HL - LH,
              HH_LH = HH - LH,
              HH_HL = HH - HL) 

# summarize mean differences
dat <- dat %>% 
  group_by(depth, microsite) %>% 
  summarize_at(vars(LH_LL, HL_LL, HH_LL, HL_LH, HH_LH, HH_HL), funs(mean, lwr, upr))

# arrange cols a bit better
dat <- select(dat, depth, microsite,
              LH_LL_mean, LH_LL_lwr, LH_LL_upr,
              HL_LL_mean, HL_LL_lwr, HL_LL_upr,
              HH_LL_mean, HH_LL_lwr, HH_LL_upr,
              HL_LH_mean, HL_LH_lwr, HL_LH_upr,
              HH_LH_mean, HH_LH_lwr, HH_LH_upr,
              HH_HL_mean, HH_HL_lwr, HH_HL_upr)

knitr::kable(dat, digits=2)

```


## Fire severity differences within depth

```{r}

dat <- melt(M.predicted, 
            id.vars = c("severity", "depth", "microsite"), 
            measure.vars = predict.cols) %>%
  mutate(value = 10^value)

dat <- dcast(dat, 
             depth + microsite + variable ~ severity, 
             value.var = "value")

# calculate pairwise differences between fire frequencies within depth within region
dat <- dat %>%
  group_by(depth) %>%
  mutate(LH_LL = LH - LL,
         HL_LL = HL - LL,
         HH_LL = HH - LL,
         HL_LH = HL - LH,
         HH_LH = HH - LH,
         HH_HL = HH - HL)

# summarize differences
deltas <- dat %>% 
  summarize_at(vars(LH_LL, HL_LL, HH_LL, HL_LH, HH_LH, HH_HL), funs(mean, lwr, upr))

# arrange cols
deltas <- deltas %>%
  select(depth,
         LH_LL_mean, LH_LL_lwr, LH_LL_upr,
         HL_LL_mean, HL_LL_lwr, HL_LL_upr,
         HH_LL_mean, HH_LL_lwr, HH_LL_upr,
         HL_LH_mean, HL_LH_lwr, HL_LH_upr,
         HH_LH_mean, HH_LH_lwr, HH_LH_upr,
         HH_HL_mean, HH_HL_lwr, HH_HL_upr)

knitr::kable(deltas, digits=2)

```


## Fire severity differences

```{r}

dat <- melt(M.predicted, 
            id.vars = c("severity", "depth", "microsite"), 
            measure.vars = predict.cols) %>%
  mutate(value = 10^value)

dat <- dcast(dat, 
             depth + microsite + variable ~ severity, 
             value.var = "value")

# calculate pairwise differences
dat <- mutate(dat,
              LH_LL = LH - LL,
              HL_LL = HL - LL,
              HH_LL = HH - LL,
              HL_LH = HL - LH,
              HH_LH = HH - LH,
              HH_HL = HH - HL)

# summarize mean differences
deltas <- dat %>% 
  group_by(depth) %>%
  summarize_at(vars(LH_LL, HL_LL, HH_LL, HL_LH, HH_LH, HH_HL), funs(mean, lwr, upr))

# munge into a more compact format
deltas <- deltas %>%
  melt(id.vars="depth") %>%
  mutate(severities = str_sub(variable, 1, 5),
         stat = str_extract(variable, "[a-z]+$"))

deltas <- dcast(deltas, severities + depth ~ stat) %>%
  select(severities, depth, mean, lwr, upr)

knitr::kable(deltas, digits=2)

```


## Separate model to estimate ratio of %RPC to %total Carbon

Graph for appendix B of paper. Note that depth labels have been changed to '0-5cm' and '5-15cm' in this graph to address reviewer comments for the paper.

```{r}

modelTxt <- "model {
  for (i in 1:N) {
    for (j in 1:Ndigest) {
      RPC[i, j] ~ dt(ytrue[i], sd.digest^(-2), nu)
    }
    
    prpc[i] <- ytrue[i] / Ctotal[i]

    ytrue[i] ~ dnorm(0, 1e-3)
  }
  
  sd.digest ~ dunif(0, 10)
  
  nuMinusOne ~ dexp(1 / 29.0)
  nu <- nuMinusOne + 1
}"


zz <- textConnection(modelTxt)

model <- jags.model(zz,
                    
                    data = list(
                      N = nrow(yobs),
                      Ndigest = ncol(yobs) - 1,
                      RPC = 10^(yobs[, -1]),
                      Ctotal = DAT.thinned$predigestPercentC
                    ),
                    
                    inits = function(...) {
                      list(
                        ytrue = rep(0, nrow(yobs)),
                        sd.digest = runif(1, 0, 1),
                        nuMinusOne = runif(1, 1, 20))
                    }, 
                    
                    n.chains = 1)

close(zz)

update(model, 1000)

sims <- coda.samples(model, 
                     variable.names = c("nu", "prpc"), 
                     n.iter = 10000)

# Transpose so model iterations are columns. Discard shape parameter samples (col 1).
m <- t(as.matrix(sims[[1]][, -1]))


dat <- cbind(DAT.thinned, m) %>%
  
  tidyr::gather(iter, value, -c(!!colnames(DAT.thinned))) %>%
  
  group_by(depth, microsite, severity) %>%
  
  summarize(lwr = lwr(value),
            upr = upr(value),
            mean = mean(value))


ggplot(data = dat) +
  geom_errorbar(aes(x = severity, group = microsite, ymin = lwr, ymax = upr),
                width = 0.5,
                position = position_dodge(width = 0.7)) +
  
  geom_point(aes(x = severity, y = mean, shape = microsite), 
             size = 3,
             position = position_dodge(width = 0.7)) +
  
  scale_shape(name = "micro-site", 
              breaks = c("O", "R", "S"),
              labels = c("open", "rough", "smooth")) +
  
  labs(x = "Fire intensity", 
       y = expression("Ratio %RPC to %"*C[Tot])) +
  
  facet_wrap(~ depth, ncol = 1,
             labeller = as_labeller(c('0-5cm' = "0-5cm", '6-15cm' = "5-15cm")))


```

